\bf Title here

### 2020-07-13 paper/paper.Rmd
more notation on the searching space and kol_cdf index

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their intial paper on projection pursuit in 1974 that "the  technique  use  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied to more general to help practitioner to unveil the black-box in randomised iteartive (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be a data matrix, an n-d projection can be seen as a linear transformation: $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times p}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis  Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar value $I$, this function is commonly known as the projection pursuit index (PPI) function, or index function and can be used to measure the "interestingness" of the projection. A number of indice functions have been proposed in the literature and this includes Legendre index [@friedman1974projection], hermite index [@hall1989polynomial], natural hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection]. [Any literature on holes index? xxx]

@@

Below we will present several examples to diagnose different aspects of the projection pursuit optimisation. We will present:

@@

3) a self-contained example to understand how to optimise a noisy and complex index function.

@@

@@

The largest difficulties of exploring searching points is its unknown number of observations per `tries`. Mapping `id` on the x-axis will leave the `tries` with few observations a small space in the plot, while those `tries` with large number of search points (usually towards the end) occupy the vast majority of the plot space. This motivates to summarise the points in each `tries`. At each iteration, rather than knowing the index value of *every* points, we are more interested to know a general summary of all the points and more importantly, the point with the largest `index_val` since it prescribes the geodesic interpolation and future searches.

*Example: exploring searching points* The larger dataset is used for 2D projection with `search_better` and `max.tries = 500`. In Figure \ref{points}, the searching points with `id` and `tries` on the x-axis, colored by `tries`. Label at the bottom indicates the number of observations in each tries and facilitates the choice of cutoff to switch from point geometry to boxplot geometry (`cutoff = 15`). The line geometry suggests the largest improvement happens at `tries = 5`.

@@

#### A more complex example: adding an interruption in the interpolation stage

@@

*Example: Interruption* This examples uses `search_better` for a 2D projection on the larger dataset using the `holes` index. In the interpolation stage, continuous frames will be constructed to bridge the project on the current basis to the target basis.  The target basis will become the current basis in the next iteration and the searching stage continues to find the next target basis. In the left panel of Figure \ref{interruption}, we observe that  there are interpolating bases with higher index values than the target basis and these bases could be used to search for new basis in the next searching stage.

@@

Exploring the value of an objective function is cna be a simple task. Because the iterator has a time series ordering feature, it goes onto the x-axis as convention. The value of hte objective function goes to the y-axis. Other parameter of interest can be represented using color to present more information. [This is the case for our purpose with method, but when the level goes higher or the parameter is quantitative, color may not be a good option. ]

An option to explore of the searching space is to explore a reduced projection 2D space via principle component analysis or other dimensional reduction methods. In this way, the first two principle components will take up the x and y axis and all the other information will be mapped using other aesthetic attribution. [need to think further about how different type of variables can be shown]. While explore the reduced space is an initial attempt to understand the searching space, there are existing technology for rotating a higher dimensional space for visualisation. Geozoo is an option. It generates random points on the high dimensional space and we can overlay it with the points on the optimisation path to visualise the spread of it on the high-D sphere.

@@

The parameter to explore can be a vector or matrix instead of scalar, for example, the projection basis in tour. This imposes difficulties in visualisation since humans are bounded to perceive at most three dimensions in a plot. Thus, principal component analysis is used to reduce the dimension of projection bases and the first two principal components are mapped to the x and y axis of the plot. Additional variable of interest could be mapped to the color aesthetics for exploration.

@@

holes_1d_geo %>% explore_proj_pca(col = info) + 
  theme(legend.position = "bottom")

@@

*Example: Noisy index function* The interpolation path of holes index, as seen in Figure \ref{interruption}, is smooth, while this may not be the case for more complicated index functions. `kol_cdf` index, an index function based on Kolmogorov test, compares the difference between a projection matrix and a randomly generated normal distribution based on cumulated distribution function (CDF). Several diagnostic visualisations below show the characteristic of this index function.

Figure \ref{kol-cdf} compares the tracing plot of the interpolating points for `search_geodesic` and `search_better` on 1D projection for `kol_cdf` index and the interpolation path shows a zig-zag pattern. Polishing step has done much more work to reach the final index value for `search_geodesic` than `search_better` and this indicates that noisy index's favour of a random search method than ascent method [the order of the examples needs to be rearranged since I haven't introduced polishing here. xxx].

@@

The second example for `kol_cdf` index uses 1D projection on the larger dataset. Since there are two informative variables and the projection is 1D, there are two local maximum  when the projection matrix at $[0, 1, 0, 0, 0, 0]$ and $[0, 0, 1 ,0, 0, 0]$. As in Figure \ref{1d-2var-different-seeds}, different local maximum can be found using `search_better` with different seeds while `search_better_random` can always find the global maximum, as in Figure \ref{1d-2var-better-random}, although at a high cost of number of points evaluated.

@@

>>>>

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their intial paper on projection pursuit in 1974 that "the  technique  use  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied to more general to help practitioner to unveil the black-box in randomised iteartive (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, bbm}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be a data matrix, an n-d projection can be seen as a linear transformation: $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis  Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar value $I$, this function is commonly known as the projection pursuit index (PPI) function, or index function and can be used to measure the "interestingness" of the projection. A number of indice functions have been proposed in the literature and this includes Legendre index [@friedman1974projection], hermite index [@hall1989polynomial], natural hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection]. [Any literature on holes index? xxx]

@@

Below we will present several examples of diagnosing different aspects of the projection pursuit optimisation. We will present:

@@

3) a self-contained example to optimise a complex index function.

@@

It is a convention to map the time index of a time series on the x-axis and value to the y-axis. In our diagnostic plots, it is to map the iterator on the x-axis and index value to the y-axis. The next question is to ask, given four indices being defined in the global object, which one should be used? *It depends on the *.

As the searching goes towards the end, it is easy to have hundreds of candidate bases in each iteration with very close index value. When plotting, these iteration will occupy the vast majority of the plot space. This motivates to summarise the points in each iteration (`tries`). At each iteration, rather than knowing the index value of *every* points, we are more interested to have a general summary of all the points and more importantly, the point with the largest `index_val` since it prescribes the next geodesic interpolation and future searches.

Using the notation we defined earlier, the plot can be defined as plotting $j$ on the x-axis and the five point summary of the following groups on the y-axis:

$$\mathbbm{1}(I_{j = j^*})$$

@@

The larger dataset is used for 2D projection with `search_better` and `max.tries = 500`. In Figure \ref{points}, the searching points with `id` and `tries` on the x-axis, colored by `tries`. Label at the bottom indicates the number of observations in each tries and facilitates the choice of cutoff to switch from point geometry to boxplot geometry (`cutoff = 15`). The line geometry suggests the largest improvement happens at `tries = 5`.

@@

The number of interpolation points is much managable than the searching points and always shows a pattern of gradual increase in the index value. Hence a line plot with interpolating order on the x-axis will be desirable. Using notation, the x-axis is a sequence $$x = \{1, 2, \cdots, k_1, \cdots, \sum_{i=1}^Jk_i\}$$ and y-axis are the index values defined by $y = I_{x}$.

#### A more complex example: Interruption

@@

This examples uses `search_better` for a 2D projection on the larger dataset using the `holes` index. In the interpolation stage, continuous frames will be constructed to bridge the project on the current basis to the target basis.  The target basis will become the current basis in the next iteration and the searching stage continues to find the next target basis. In the left panel of Figure \ref{interruption}, we observe that  there are interpolating bases with higher index values than the target basis and these bases could be used to search for new basis in the next searching stage.

@@

@@

In projection pursuit, the projection bases $\mathbf{A}_{p \times d}$ are usually of dimension $p \times d$ and hence can't be visualised in a 2D plot. An option to explore the searching space of these bases is to explore a reduced space via principal component analysis (PCA). We first flat each projection bases $\mathbf{A}$ to a row vector and stack them by rows. This gives a matrix of dimension $n$ by $(p*d)$. Centering and scaling are usually done before performing PCA and we denotes the pre-processed matrix as $\tilde{\mathbf{A}}_{n \times (p \times d)}$. By definition, the principal components maximise the variance of the projection bases in the reduced space. Using lagrange multipler, this is equivalent to find the eigenvector of the variance-covariance matrix: $\Sigma = \tilde{\mathbf{A}}^T\tilde{\mathbf{A}}$. Decomposing $\tilde{\mathbf{A}}$ using singular value decomposition (SVD) gives $$\tilde{\mathbf{A}} = \mathbf{U} \mathbf{\lambda} \mathbf{V}$$ and hence the variance-covariance matrix

$$\Sigma = \tilde{\mathbf{A}}^T\tilde{\mathbf{A}} = (\mathbf{V}^T \mathbf{\lambda}^T \mathbf{U}^T) (\mathbf{U} \mathbf{\lambda} \mathbf{V}) = \mathbf{V}^T \mathbf{\lambda}^2 \mathbf{V}$$

has eigenvector $\mathbf{V} = [\mathbf{v}_1, \mathbf{v}_2, \cdots, \mathbf{v}_{p\times d}]$. The projection mapping to the x- and y-axis uses the first two principal components and hence $$\tilde{\mathbf{A}} \mathbf{v}_1 \text{ and } \tilde{\mathbf{A}} \mathbf{v}_2$$

The resulting two vectors are of dimension $n \times 1$, which is compatible with the dimension of the global object. THis means all the iterators and interesting other variables defined in the global object can be binded with the these two projections and can be mapped to other aesthetics (i.e. color).

In this way, the first two principle components will take up the x and y axis and all the other information will be mapped using other aesthetic attribution. [need to think further about how different type of variables can be shown]. While explore the reduced space is an initial attempt to understand the searching space, there are existing technology for rotating a higher dimensional space for visualisation. Geozoo is an option. It generates random points on the high dimensional space and we can overlay it with the points on the optimisation path to visualise the spread of it on the high-D sphere.

@@

holes_1d_geo %>% explore_proj_pca(col = info)

@@

The interpolation path of holes index, as seen in Figure \ref{interruption}, is smooth, while this may not be the case for more complicated index functions. `kol_cdf` index, an 1D projection index function based on Kolmogorov test, compares the difference between the 1D projected data, $\mathbf{P}_{n \times 1}$ and a randomly generated normal distribution, $y_n$ based on the empirical cumulated distribution function (ECDF). Denotes the ECDF function as $F(u)$ with subscript incidating the variable,  the Kolmogorove statistics defined by

$$\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]$$

can be seen as a function of the projection matrix $\mathbf{A}_{p \times 1}$ and hence a valid index function.

#### Explore index value

Figure \ref{kol-cdf} compares the tracing plot of the interpolating points when using different optimisation algorithms: `search_geodesic` and `search_better`. One can observe that

- The index value of `kol_cdf` index is much smaller than that of holes index
- The link of index values from interpolation bases  are no longer smooth
- Both algorithms reach a similar final index value after polishing

Polishing step has done much more work to find the final index value ub `search_geodesic` than `search_better` and this indicates `kol_cdf` function favours of a random search method than ascent method.

@@

Now we enlarge the dataset to include two informative variables: `x2` and `x3` and remain 1D projection. In this case, two local maximum appear when projection matrix being $[0, 1, 0, 0, 0, 0]$ and $[0, 0, 1 ,0, 0, 0]$.

Using different seeds in `search_better` allows us to find both local maximum as in Figure \ref{1d-2var-different-seeds}. Comparing the maximum of both, we can see that the global maximum happens when `x2` is found. It is natural to ask then if there is an algorithm that can find the global maximum without trying on different seeds? `search_better_random` manages to do it via a metropolis-hasting random search as shown in Figure \ref{1d-2var-better-random}, although at a higher cost of number of points to evaluate.

@@

#### Explore searching space

We can also plot the searching points of all the three algorithms in the searching space and explore their relative position against each other using principal components. As shown in Figure \ref{1d-2var-explore-proj-pca}, the bases from better1 and better2 only search a proportion of the searching space while better_random produces a more exhausive search. The large overlapping of better1 and better_random is explained by the fact that both algorithms finds x2 in the end.

```{r 1d-2var-explore-proj-pca, fig.cap = "\\label{1d-2var-explore-proj-pca} The projected projection basis using principal components. The bases from better1 and better2 only search a proportion of the searching space while better_random produces a more exhausive search. The large overlapping of better1 and better_random is explained by the fact that both algorithms finds x2 in the end."}
interp1 <- kol_cdf_1d_2var_better 
interp2 <- kol_cdf_1d_2var_better_2
interp3 <- kol_cdf_1d_2var_better_random

bind_rows(interp1, interp2, interp3) %>% 
  mutate(algorithm = c(rep("better1", nrow(interp1)), 
                       rep("better2", nrow(interp2)), 
                       rep("better_random", nrow(interp3))),
         algorithm = fct_relevel(algorithm, c("better1", "better2", "better_random"))) %>% 
  explore_proj_pca(col = algorithm) + 
  scale_color_brewer(palette = "Dark2")

### 2020-07-13 paper/paper.tex
more notation on the searching space and kol_cdf index

%% load any required packages here
\usepackage{amssymb, amsmath, mathtools} \usepackage[ruled,vlined, linesnumbered]{algorithm2e}

@@

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\) be a
data matrix, an n-d projection can be seen as a linear transformation:
\(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times p}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or index function and can be used to measure the
``interestingness'' of the projection. A number of indice functions have
been proposed in the literature and this includes Legendre index
\citep{friedman1974projection}, hermite index
\citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}. {[}Any literature on holes index?
xxx{]}

@@

Below we will present several examples to diagnose different aspects of
the projection pursuit optimisation. We will present:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  static plots to explore the index value,
\item
  animated plots to explore the projection basis and,
\item
  a self-contained example to understand how to optimise a noisy and
  complex index function.
\end{enumerate}

@@

@@

The largest difficulties of exploring searching points is its unknown
number of observations per \texttt{tries}. Mapping \texttt{id} on the
x-axis will leave the \texttt{tries} with few observations a small space
in the plot, while those \texttt{tries} with large number of search
points (usually towards the end) occupy the vast majority of the plot
space. This motivates to summarise the points in each \texttt{tries}. At
each iteration, rather than knowing the index value of \emph{every}
points, we are more interested to know a general summary of all the
points and more importantly, the point with the largest
\texttt{index\_val} since it prescribes the geodesic interpolation and
future searches.

\emph{Example: exploring searching points} The larger dataset is used
for 2D projection with \texttt{search\_better} and
\texttt{max.tries\ =\ 500}. In Figure \ref{points}, the searching points
with \texttt{id} and \texttt{tries} on the x-axis, colored by
\texttt{tries}. Label at the bottom indicates the number of observations
in each tries and facilitates the choice of cutoff to switch from point
geometry to boxplot geometry (\texttt{cutoff\ =\ 15}). The line geometry
suggests the largest improvement happens at \texttt{tries\ =\ 5}.

@@

\hypertarget{a-more-complex-example-adding-an-interruption-in-the-interpolation-stage}{%
\paragraph{A more complex example: adding an interruption in the
interpolation
stage}\label{a-more-complex-example-adding-an-interruption-in-the-interpolation-stage}}

\emph{Example: Interruption} This examples uses \texttt{search\_better}
for a 2D projection on the larger dataset using the \texttt{holes}
index. In the interpolation stage, continuous frames will be constructed
to bridge the project on the current basis to the target basis. The
target basis will become the current basis in the next iteration and the
searching stage continues to find the next target basis. In the left
panel of Figure \ref{interruption}, we observe that there are
interpolating bases with higher index values than the target basis and
these bases could be used to search for new basis in the next searching
stage.

@@

Exploring the value of an objective function is cna be a simple task.
Because the iterator has a time series ordering feature, it goes onto
the x-axis as convention. The value of hte objective function goes to
the y-axis. Other parameter of interest can be represented using color
to present more information. {[}This is the case for our purpose with
method, but when the level goes higher or the parameter is quantitative,
color may not be a good option. {]}

An option to explore of the searching space is to explore a reduced
projection 2D space via principle component analysis or other
dimensional reduction methods. In this way, the first two principle
components will take up the x and y axis and all the other information
will be mapped using other aesthetic attribution. {[}need to think
further about how different type of variables can be shown{]}. While
explore the reduced space is an initial attempt to understand the
searching space, there are existing technology for rotating a higher
dimensional space for visualisation. Geozoo is an option. It generates
random points on the high dimensional space and we can overlay it with
the points on the optimisation path to visualise the spread of it on the
high-D sphere.

@@

The parameter to explore can be a vector or matrix instead of scalar,
for example, the projection basis in tour. This imposes difficulties in
visualisation since humans are bounded to perceive at most three
dimensions in a plot. Thus, principal component analysis is used to
reduce the dimension of projection bases and the first two principal
components are mapped to the x and y axis of the plot. Additional
variable of interest could be mapped to the color aesthetics for
exploration.

@@

\emph{Example: Noisy index function} The interpolation path of holes
index, as seen in Figure \ref{interruption}, is smooth, while this may
not be the case for more complicated index functions. \texttt{kol\_cdf}
index, an index function based on Kolmogorov test, compares the
difference between a projection matrix and a randomly generated normal
distribution based on cumulated distribution function (CDF). Several
diagnostic visualisations below show the characteristic of this index
function.

Figure \ref{kol-cdf} compares the tracing plot of the interpolating
points for \texttt{search\_geodesic} and \texttt{search\_better} on 1D
projection for \texttt{kol\_cdf} index and the interpolation path shows
a zig-zag pattern. Polishing step has done much more work to reach the
final index value for \texttt{search\_geodesic} than
\texttt{search\_better} and this indicates that noisy index's favour of
a random search method than ascent method {[}the order of the examples
needs to be rearranged since I haven't introduced polishing here.
xxx{]}.

@@

The second example for \texttt{kol\_cdf} index uses 1D projection on the
larger dataset. Since there are two informative variables and the
projection is 1D, there are two local maximum when the projection matrix
at \([0, 1, 0, 0, 0, 0]\) and \([0, 0, 1 ,0, 0, 0]\). As in Figure
\ref{1d-2var-different-seeds}, different local maximum can be found
using \texttt{search\_better} with different seeds while
\texttt{search\_better\_random} can always find the global maximum, as
in Figure \ref{1d-2var-better-random}, although at a high cost of number
of points evaluated.

@@

>>>>

%% load any required packages here
\usepackage{amssymb, amsmath, mathtools, bbm} \usepackage[ruled,vlined, linesnumbered]{algorithm2e}

@@

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. Let \(\mathbf{X}_{n \times p}\) be a
data matrix, an n-d projection can be seen as a linear transformation:
\(T: \mathbb{R}^p \mapsto \mathbb{R}^d\) defined by
\(\mathbf{P} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{P}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis Define
\(f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}\) be an index function
that maps the projection basis \(\mathbf{A}\) onto a scalar value \(I\),
this function is commonly known as the projection pursuit index (PPI)
function, or index function and can be used to measure the
``interestingness'' of the projection. A number of indice functions have
been proposed in the literature and this includes Legendre index
\citep{friedman1974projection}, hermite index
\citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}. {[}Any literature on holes index?
xxx{]}

@@

Below we will present several examples of diagnosing different aspects
of the projection pursuit optimisation. We will present:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  static plots to explore the index value,
\item
  animated plots to explore the projection basis and,
\item
  a self-contained example to optimise a complex index function.
\end{enumerate}

@@

It is a convention to map the time index of a time series on the x-axis
and value to the y-axis. In our diagnostic plots, it is to map the
iterator on the x-axis and index value to the y-axis. The next question
is to ask, given four indices being defined in the global object, which
one should be used? \emph{It depends on the }.

As the searching goes towards the end, it is easy to have hundreds of
candidate bases in each iteration with very close index value. When
plotting, these iteration will occupy the vast majority of the plot
space. This motivates to summarise the points in each iteration
(\texttt{tries}). At each iteration, rather than knowing the index value
of \emph{every} points, we are more interested to have a general summary
of all the points and more importantly, the point with the largest
\texttt{index\_val} since it prescribes the next geodesic interpolation
and future searches.

Using the notation we defined earlier, the plot can be defined as
plotting \(j\) on the x-axis and the five point summary of the following
groups on the y-axis:

\[\mathbbm{1}(I_{j = j^*})\]

@@

The larger dataset is used for 2D projection with
\texttt{search\_better} and \texttt{max.tries\ =\ 500}. In Figure
\ref{points}, the searching points with \texttt{id} and \texttt{tries}
on the x-axis, colored by \texttt{tries}. Label at the bottom indicates
the number of observations in each tries and facilitates the choice of
cutoff to switch from point geometry to boxplot geometry
(\texttt{cutoff\ =\ 15}). The line geometry suggests the largest
improvement happens at \texttt{tries\ =\ 5}.

@@

The number of interpolation points is much managable than the searching
points and always shows a pattern of gradual increase in the index
value. Hence a line plot with interpolating order on the x-axis will be
desirable. Using notation, the x-axis is a sequence
\[x = \{1, 2, \cdots, k_1, \cdots, \sum_{i=1}^Jk_i\}\] and y-axis are
the index values defined by \(y = I_{x}\).

\hypertarget{a-more-complex-example-interruption}{%
\paragraph{A more complex example:
Interruption}\label{a-more-complex-example-interruption}}

This examples uses \texttt{search\_better} for a 2D projection on the
larger dataset using the \texttt{holes} index. In the interpolation
stage, continuous frames will be constructed to bridge the project on
the current basis to the target basis. The target basis will become the
current basis in the next iteration and the searching stage continues to
find the next target basis. In the left panel of Figure
\ref{interruption}, we observe that there are interpolating bases with
higher index values than the target basis and these bases could be used
to search for new basis in the next searching stage.

@@

@@

In projection pursuit, the projection bases \(\mathbf{A}_{p \times d}\)
are usually of dimension \(p \times d\) and hence can't be visualised in
a 2D plot. An option to explore the searching space of these bases is to
explore a reduced space via principal component analysis (PCA). We first
flat each projection bases \(\mathbf{A}\) to a row vector and stack them
by rows. This gives a matrix of dimension \(n\) by \((p*d)\). Centering
and scaling are usually done before performing PCA and we denotes the
pre-processed matrix as \(\tilde{\mathbf{A}}_{n \times (p \times d)}\).
By definition, the principal components maximise the variance of the
projection bases in the reduced space. Using lagrange multipler, this is
equivalent to find the eigenvector of the variance-covariance matrix:
\(\Sigma = \tilde{\mathbf{A}}^T\tilde{\mathbf{A}}\). Decomposing
\(\tilde{\mathbf{A}}\) using singular value decomposition (SVD) gives
\[\tilde{\mathbf{A}} = \mathbf{U} \mathbf{\lambda} \mathbf{V}\] and
hence the variance-covariance matrix

\[\Sigma = \tilde{\mathbf{A}}^T\tilde{\mathbf{A}} = (\mathbf{V}^T \mathbf{\lambda}^T \mathbf{U}^T) (\mathbf{U} \mathbf{\lambda} \mathbf{V}) = \mathbf{V}^T \mathbf{\lambda}^2 \mathbf{V}\]

has eigenvector
\(\mathbf{V} = [\mathbf{v}_1, \mathbf{v}_2, \cdots, \mathbf{v}_{p\times d}]\).
The projection mapping to the x- and y-axis uses the first two principal
components and hence
\[\tilde{\mathbf{A}} \mathbf{v}_1 \text{ and } \tilde{\mathbf{A}} \mathbf{v}_2\]

The resulting two vectors are of dimension \(n \times 1\), which is
compatible with the dimension of the global object. THis means all the
iterators and interesting other variables defined in the global object
can be binded with the these two projections and can be mapped to other
aesthetics (i.e.~color).

In this way, the first two principle components will take up the x and y
axis and all the other information will be mapped using other aesthetic
attribution. {[}need to think further about how different type of
variables can be shown{]}. While explore the reduced space is an initial
attempt to understand the searching space, there are existing technology
for rotating a higher dimensional space for visualisation. Geozoo is an
option. It generates random points on the high dimensional space and we
can overlay it with the points on the optimisation path to visualise the
spread of it on the high-D sphere.

@@

The interpolation path of holes index, as seen in Figure
\ref{interruption}, is smooth, while this may not be the case for more
complicated index functions. \texttt{kol\_cdf} index, an 1D projection
index function based on Kolmogorov test, compares the difference between
the 1D projected data, \(\mathbf{P}_{n \times 1}\) and a randomly
generated normal distribution, \(y_n\) based on the empirical cumulated
distribution function (ECDF). Denotes the ECDF function as \(F(u)\) with
subscript incidating the variable, the Kolmogorove statistics defined by

\[\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]\]

can be seen as a function of the projection matrix
\(\mathbf{A}_{p \times 1}\) and hence a valid index function.

\hypertarget{explore-index-value}{%
\paragraph{Explore index value}\label{explore-index-value}}

Figure \ref{kol-cdf} compares the tracing plot of the interpolating
points when using different optimisation algorithms:
\texttt{search\_geodesic} and \texttt{search\_better}. One can observe
that

\begin{itemize}
\tightlist
\item
  The index value of \texttt{kol\_cdf} index is much smaller than that
  of holes index
\item
  The link of index values from interpolation bases are no longer smooth
\item
  Both algorithms reach a similar final index value after polishing
\end{itemize}

Polishing step has done much more work to find the final index value ub
\texttt{search\_geodesic} than \texttt{search\_better} and this
indicates \texttt{kol\_cdf} function favours of a random search method
than ascent method.

@@

Now we enlarge the dataset to include two informative variables:
\texttt{x2} and \texttt{x3} and remain 1D projection. In this case, two
local maximum appear when projection matrix being \([0, 1, 0, 0, 0, 0]\)
and \([0, 0, 1 ,0, 0, 0]\).

Using different seeds in \texttt{search\_better} allows us to find both
local maximum as in Figure \ref{1d-2var-different-seeds}. Comparing the
maximum of both, we can see that the global maximum happens when
\texttt{x2} is found. It is natural to ask then if there is an algorithm
that can find the global maximum without trying on different seeds?
\texttt{search\_better\_random} manages to do it via a
metropolis-hasting random search as shown in Figure
\ref{1d-2var-better-random}, although at a higher cost of number of
points to evaluate.

@@

\hypertarget{explore-searching-space}{%
\paragraph{Explore searching space}\label{explore-searching-space}}

We can also plot the searching points of all the three algorithms in the
searching space and explore their relative position against each other
using principal components. As shown in Figure
\ref{1d-2var-explore-proj-pca}, the bases from better1 and better2 only
search a proportion of the searching space while better\_random produces
a more exhausive search. The large overlapping of better1 and
better\_random is explained by the fact that both algorithms finds x2 in
the end.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/1d-2var-explore-proj-pca-1.pdf}
\caption{\label{1d-2var-explore-proj-pca} The projected projection basis
using principal components. The bases from better1 and better2 only
search a proportion of the searching space while better\_random produces
a more exhausive search. The large overlapping of better1 and
better\_random is explained by the fact that both algorithms finds x2 in
the end.}
\end{figure}
